Improving the portuguese hyphenation rules

### 2024-06-14 article/updatingrules.tex
updates

\section{Updating the rules}
We now begin incorporating the rules proposed in the previous section and
appraise the results of each one. During this process, we will consider the
necessity of updating those rules, including exceptions, or even discarding the
proposed rules if needed.

@@

% cat ../data/default.TeX.pt-br.patterns <(echo -e "p2si\np2sí\n.g2no\n.g2nó\nt2c\np2neu\n.t2m\n.p2t\nsu2b3r\nsu2b3l") > /tmp/patts
% ./gocreateresultfile.sh ../data/hyphenations6.dic /tmp/patts > /tmp/hyphenations6result_001.dic
% ./resultfilemetrics.sh -f /tmp/hyphenations6result_001.dic
% correct: 15552; wrong: 6 (6); missing: 286 (285,1)
%
% compare with the results of the default rules:
% diff -y --suppress-common-lines <(sort ../data/hyphenations6result-default-patterns.dic) <(sort /tmp/hyphenations6result_001.dic)
% 
% ./gohyphenfull ../data/default.TeX.pt-br.patterns psiquiátrico
%
% cat ../data/default.TeX.pt-br.patterns <(echo -e ".p2si\n.p2sí\n.g2no\n.g2nó\nt2c\np2neu\n.t2m\n.p2t\nsu2b3r\nsu2b3l") > /tmp/patts
% ./gocreateresultfile.sh ../data/hyphenations6.dic /tmp/patts > /tmp/hyphenations6result_001.dic
% ./resultfilemetrics.sh -f /tmp/hyphenations6result_001.dic
% correct: 15562; wrong: 3 (3); missing: 279 (278,1)

After introducing the rules: \texttt{.p2si}, \texttt{.p2sí}, \texttt{.g2no}, \texttt{.g2nó}, \texttt{t2c}, 
\texttt{p2neu}, \texttt{.t2m}, \texttt{.p2t}, \texttt{su2b3r}, and \texttt{su2b3l}, 
the number of incorrect hyphenations decreased to 3, and the number of missing hyphenations decreased to 278.
At first sight, we expected the number of incorrect hyphenations to drop to zero. However, 
the introduction of rule \texttt{su2b3l} introduced hyphenation errors in \emph{su-b.li*me}, \emph{su-b.li*ma*ção} and \emph{su-b.li*ma*do}.

\noindent\begin{minipage}{\linewidth}
\begin{lstlisting}[language={}, caption={Hyphenation of the word \emph{sublime} after the indroduction of rule \texttt{su2b3l}. 
This rule introduces a wrong hyphenation point, demanding an exception rule to fix it.}, label=sublimehyphenation]
   .   s   u   b   l   i   m   e   .
     1   0   0   |   |   |   |   |    1su
     0   0   2   3   0   |   |   |    su2b3l
     |   |   1   2   0   |   |   |    1b2l
     |   |   |   1   0   0   |   |    1li
     |   |   |   |   |   1   0   0    1me
max: 1   0   2   3   0   1   0   0
final: s   u   b - l   i - m   e
\end{lstlisting}
\end{minipage}
\noindent{}By inserting the rule \texttt{.su3b4li}, we fixed the hyphenation errors caused by the introduction of the rule \texttt{su2b3l},
as illustrated in \Cref{sublimehyphenation}. This adjustment resulted in zero hyphenation errors and 276 missing hyphens.
% cat ../data/default.TeX.pt-br.patterns <(echo -e ".p2si\n.p2sí\n.g2no\n.g2nó\nt2c\np2neu\n.t2m\n.p2t\nsu2b3r\nsu2b3l\n.su3b4li") > /tmp/patts
% ./gocreateresultfile.sh ../data/hyphenations6.dic /tmp/patts > /tmp/hyphenations6result_001.dic
% ./resultfilemetrics.sh -f /tmp/hyphenations6result_001.dic
% correct: 15565; wrong: 0 (); missing: 276 (275,1)

@@

The next step is to introduce the rules o address the missing hyphenation points. These rules are: 
\texttt{a1ir.}, \texttt{u1ir.}, \texttt{i3â}, \texttt{i3ê}, \texttt{i3ô}, \texttt{1qu}, \texttt{1vô}, \texttt{1lô}, \texttt{1cô}, \texttt{1gô}, \texttt{1bô}, 
\texttt{1tô}, \texttt{1rô}, \texttt{1pô}, \texttt{a1é}, \texttt{a1í}, \texttt{a1ó}, \texttt{a1ú}, \texttt{e1á}, \texttt{e1â}, \texttt{e1ã}, 
\texttt{e1é}, \texttt{e1ê}, \texttt{e1í}, \texttt{e1ó}, \texttt{é1o}, \texttt{e1ú}, \texttt{i1á}, \texttt{i1ã}, \texttt{í1a}, \texttt{i1é}, 
\texttt{i1í}, \texttt{i1ó}, \texttt{í1o}, \texttt{i1u}, \texttt{i1ú}, \texttt{o1á}, \texttt{o1é}, \texttt{o1í}, \texttt{o1ó}, \texttt{u1á}, \texttt{u1ã}, 
\texttt{u1í}, \texttt{ú1o}.
% \texttt{1qu2á}, \texttt{1qu2é}, \texttt{1qu2ê}, \texttt{1qu2í}, \texttt{que2i}, \texttt{1qu4ã}, \texttt{1qu4ó}, \texttt{1qu4â}
% \texttt{1gu2é}, \texttt{1gu2ê}, \texttt{1gu2í} 
% cat ../data/default.TeX.pt-br.patterns <(echo ".p2si, .p2sí, .g2no, .g2nó, t2c, p2neu, .t2m, .p2t, su2b3r, su2b3l, .su3b4li, a1ir., u1ir., i3â, i3ê, i3ô, 1qu, 1vô, 1lô, 1cô, 1gô, 1bô, 1tô, 1rô, 1pô, a1é, a1í, a1ó, a1ú, e1á, e1â, e1ã, e1é, e1ê, e1í, e1ó, é1o, e1ú, i1á, i1ã, í1a, i1é, i1í, i1ó, í1o, i1u, i1ú, o1á, o1é, o1í, o1ó, u1á, u1ã, u1í, ú1o" | tr -d ' ' | tr ',' '\n') > /tmp/patts2
% %%% a1â, a1ã, a1é, a1í, a1ó, a1ô, a1ú, e1á, e1â, e1ã, e1é, e1ê, e1í, e1ó, e1ô, e1ú, é1o, i1á, i1ã, i1é, i1í, i1ó, i1u, i1ú, í1a, o1á, o1ã, o1é, o1ê, o1í, o1ó, u1á, u1ã, u1é, u1ê, u1í, ú1o
% ./gocreateresultfile.sh ../data/hyphenations6.dic /tmp/patts2 > /tmp/hyphenations6result_002.dic
% ./resultfilemetrics.sh -f /tmp/hyphenations6result_002.dic
% correct: 15804; wrong: 18 (18); missing: 19 (19)
% grep "\." /tmp/hyphenations6result_002.dic | grep -oP "[^\.\*\-]{1,2}\.[^\.\*\-]{1,2}" | sort | uniq -c | sort -nr | column
% 11 qu.í          4 qu.á          1 qu.ís         1 qu.ín         1 gu.ís
After inserting these rules, the number of incorrect hyphenations increases to 18 due to the inclusion of hyphens in \emph{qu.í}, \emph{gu.í} and \emph{qu.á}. 
To overcome these errors, we need to add exception rules: \texttt{1qu2á}, \texttt{1qu2í}, and \texttt{1gu2í}.
See more on this on \cref{rulegrp_aa,rulegrp_gua}.
% cat ../data/default.TeX.pt-br.patterns <(echo ".p2si, .p2sí, .g2no, .g2nó, t2c, p2neu, .t2m, .p2t, su2b3r, su2b3l, .su3b4li, a1ir., u1ir., i3â, i3ê, i3ô, 1qu, 1vô, 1lô, 1cô, 1gô, 1bô, 1tô, 1rô, 1pô, a1é, a1í, a1ó, a1ú, e1á, e1â, e1ã, e1é, e1ê, e1í, e1ó, é1o, e1ú, i1á, i1ã, í1a, i1é, i1í, i1ó, í1o, i1u, i1ú, o1á, o1é, o1í, o1ó, u1á, u1ã, u1í, ú1o, 1qu2á, 1qu2í, 1gu2í" | tr -d ' ' | tr ',' '\n') > /tmp/patts3
% ./gocreateresultfile.sh ../data/hyphenations6.dic /tmp/patts3 > /tmp/hyphenations6result_003.dic
% ./resultfilemetrics.sh -f /tmp/hyphenations6result_003.dic
% correct: 15822; wrong: 0 (); missing: 19 (19)

We have reached a point where there is no wrong hyphenation but still there are 19 missing hyphenation. Here is the list of the missing ones:
\begin{multicols}{5}
\setlength{\columnsep}{0pt}
\setlength{\parindent}{0pt}
\emph{a-in*da \\ ra-i*nha \\ con*tri*bu-in*te \\ ra-iz \\ re-in*te*gra*ção \\ cam*pa-i*nha \\ re-in*te*grar \\ ta-i*nha \\ re-in*ci*dir \\ pa-in*ço \\ pi*xa-im \\ re-im*pres*são \\ tu-im \\ je*su-i*tis*mo \\ ben*jo-im \\ ma*lau-i*a*no \\ re-im*pri*mir}
\end{multicols}
\noindent{}Observing the instances above, 17 out of 19 involve a missing hyphen before the vowel \emph{i}. The following rules
will be used to solve those issues: \texttt{a1ind}, \texttt{a1i1nh} (see \cref{rulegrp_ain}), \texttt{e1imp} (see \cref{rulegrp_eim}), 
\texttt{e1inc}, \texttt{e1inf}, \texttt{e1ing}, \texttt{e1ins}, \texttt{e1int}, \texttt{e1inv}, (see \cref{rulegrp_ein}), 
\texttt{u1int}, \texttt{u1ind} (see \cref{rulegrp_unit}), and \texttt{u1iz.}, \texttt{a1iz.} (see \cref{rulegrp_aiz}).
% rulegrp_ain regra: a1ind, a1i1nh – ainda, rainha
% rulegrp_eim regra: e1imp – reimpresso
% rulegrp_ein regra: e1inc, e1inf, e1ing, e1ins, e1int, e1inv
% rulegrp_unit regra: u1int, u1ind -- contribuinte, excluindo
% rulegrp_aiz regra: u1iz., a1iz. -- juiz, raiz

% cat ../data/default.TeX.pt-br.patterns <(echo ".p2si, .p2sí, .g2no, .g2nó, t2c, p2neu, .t2m, .p2t, su2b3r, su2b3l, .su3b4li, a1ir., u1ir., i3â, i3ê, i3ô, 1qu, 1vô, 1lô, 1cô, 1gô, 1bô, 1tô, 1rô, 1pô, a1é, a1í, a1ó, a1ú, e1á, e1â, e1ã, e1é, e1ê, e1í, e1ó, é1o, e1ú, i1á, i1ã, í1a, i1é, i1í, i1ó, í1o, i1u, i1ú, o1á, o1é, o1í, o1ó, u1á, u1ã, u1í, ú1o, 1qu2á, 1qu2í, 1gu2í, a1ind, a1i1nh, e1imp, e1inc, e1inf, e1ing, e1ins, e1int, e1inv, u1int, u1ind, u1iz., a1iz." | tr -d ' ' | tr ',' '\n') > /tmp/patts4
% ./gocreateresultfile.sh ../data/hyphenations6.dic /tmp/patts4 > /tmp/hyphenations6result_004.dic
% ./resultfilemetrics.sh -f /tmp/hyphenations6result_004.dic
% correct: 15833; wrong: 0 (); missing: 8 (8)

>>>>

\section{Updating the rules}

In this section we will analyse the results of the default rules on our dictionaries,
find patterns, and gradually incorporate new rules to the set, appraising the result of
each one, to avoid causing more damage than good, create exception rules when necessary,
and, if necessary, discard rules. We will stick to the \emph{patgen} notation to express
a wrong, a missing, and a correct hyphenation point (`.', `-', and `*', respectively).

For the list comprised in \texttt{hyphenations6}, below is the complete list of
\NumberOfWrongSix{} words that were incorrectly hyphenated:
% wrong
\begin{multicols}{3}
\setlength{\columnsep}{0pt}
\setlength{\parindent}{0pt}
\emph{\input{/tmp/hyphenations6result-default-patterns_wrong.dic}}
\end{multicols}
In each instance, an erroneous hyphenation point was placed at the beginning of
a word, indicating a potential issue in recognizing certain prefixing morphemes
in the language.  Among those, 20 occurred in words containing the \emph{psi}
morpheme, where the algorithm erroneously hyphenated between the initial
\emph{p} and the following \emph{s}. Additionally, 3 cases involved the
starting \emph{gno} sequence, while the remaining errors were observed in the
sequences \emph{pneu} (3), \emph{tch} (1), \emph{tme} (1), \emph{pto} (1), and
\emph{sub} (1).

Those erroneous hyphenation points might be corrected by the introduction of a few rules:
\texttt{.p2si, .p2sí} (see \cref{rulegrp_psi}), \texttt{.g2no, .g2nó} (see \cref{rulegrp_gno}), 
\texttt{1p2neu} (see \cref{rulegrp_pneu}),
\texttt{t2c} (see \cref{rulegrp_tc}), \texttt{.t2m} (see \cref{rulegrp_tm}), \texttt{.p2t} (see \cref{rulegrp_pt}) 
and \texttt{su2b3r}, \texttt{su2b3l}  (see \cref{rulegrp_sub}). They are deeply 
discussed in \Cref{sec-hyph-rpatches}.

% $ ../scripts/resultfileexamples.sh -m -f ../data/hyphenations6result-default-patterns.dic | grep -o "[^-]-[^-]" | sort | uniq -c | sort -nr | column

The first 20 examples of missing hyphenations are displayed in the following list:
% missing
\begin{multicols}{3}
\setlength{\columnsep}{0pt}
\setlength{\parindent}{0pt}
\emph{\input{/tmp/hyphenations6result-default-patterns_missing.dic}
...}
\end{multicols}
It is a long list with \NumberOfMissingSix{} entries, therefore it is
unavoidable to analyse it through clusters of certain patterns. It is presented
below the counts of immediate context in which those missing hyphenations were
found:
\begin{multicols}{5}
\setlength{\columnsep}{0pt}
\setlength{\parindent}{0pt}
\emph{\input{/tmp/hyphenations6result-default-patterns_missing_context}}
\end{multicols}

% number of words correctly hyphenated using default TeX rules when analysing the words in hyphenations6.dic
%\captureshell*[\NumberOfCorrectSix]{../scripts/resultfilemetrics.sh -f ../data/hyphenations6result-default-patterns.dic -c}

% number of words with wrong hyphenations using default TeX rules when analysing the words in hyphenations6.dic
%\captureshell*[\NumberOfWrongSix]{../scripts/resultfilemetrics.sh -f ../data/hyphenations6result-default-patterns.dic -w}

% number of words with missing hyphenations using default TeX rules when analysing the words in hyphenations6.dic
%\captureshell*[\NumberOfMissingSix]{../scripts/resultfilemetrics.sh -f ../data/hyphenations6result-default-patterns.dic -m}

% tra-b.a*lho
%    ^ ^ ^
%    | | |
%    | | + -- correct (*)
%    | + ---- wrong   (.)  
%    + ------ missed  (-)

On the top of the list we see the missing hyphen in \emph{u-i}, accounting for
38 cases. Among those, 34 were originated from the pattern \texttt{u-ir} in the
end of word and 13 from \texttt{a-ir} in the end of word. 
To fix them, We will include the rules \texttt{u1ir.} and \texttt{a1ir.} (see 
\cref{rulegrp_air}). The cases that are not covered by this rule are:
\emph{tu-im}, \emph{je*su-i*tis*mo}, \emph{ma*lau-i*a*no}, and
\emph{con*tri*bu-in*te}. On \cref{rulegrp_Vi} we see how to deal with
them.

The default rules already include 17 separating rules for vowel sequences:
% vowel="[aeiouáãàâéêíóõôú]"; grep "$vowel[13579]$vowel" ../data/default.TeX.pt-br.patterns | sed ':a;N;$!ba;s/\n/ \\\\ /g'
\begin{multicols}{5}
\setlength{\columnsep}{0pt}
\setlength{\parindent}{0pt}
\texttt{a3a \\ a3e \\ a3o \\ e3a \\ e3e \\ e3o \\ i3a \\ i3e \\ i3i \\ i3o \\ i3â \\ i3ê \\ i3ô \\ o3a \\ o3e \\ o3o \\ u3a \\ u3e \\ u3o \\ u3u}
\end{multicols}
\noindent{}but there are just 3 rules to hyphenate between vowels when one has a diacritic: \texttt{i3â}, \texttt{i3ê}, \texttt{i3ô}.
We could then add more 29 rules to account for the missing hyphenations between vowels with diacritics and also the
missing rule \texttt{i1u}:
% ../scripts/resultfileexamples.sh -m -f ../data/hyphenations6result-default-patterns.dic | ../scripts/filtersurrounding.sh -c '-' -n 1 | sort | uniq -c | sort -nr | grep "[áàâãéêíóôõú]" | awk '{print $2}' | sort | sed ':a;N;$!ba;s/\n/ \\\\ /g'
\begin{multicols}{5}
\setlength{\columnsep}{0pt}
\setlength{\parindent}{0pt}
\texttt{%
% complete set of 37 rules
%a1â \\ a1ã \\ a1é \\ a1í \\ a1ó \\ a1ô \\ a1ú \\ 
%e1á \\ e1â \\ e1ã \\ e1é \\ e1ê \\ e1í \\ e1ó \\ e1ô \\ e1ú \\ é1o \\ 
%i1á \\ i1ã \\ i1ã \\ i1é \\ i1í \\ i1ó \\ i1u \\ i1ú \\ í1a \\ 
%o1á \\ o1ã \\ o1é \\ o1ê \\ o1í \\ o1ó \\ 
%u1á \\ u1ã \\ u1é \\ u1ê \\ u1í \\ ú1o
%
% 29 rules necessary for hyphenations6 dic
a1é \\ a1í \\ a1ó \\ a1ú \\ 
e1á \\ e1â \\ e1ã \\ e1é \\ e1ê \\ e1í \\ e1ó \\ e1ú \\ é1o \\ 
i1á \\ i1ã \\ í1a \\ i1é \\ i1í \\ i1ó \\ i1u \\ i1ú \\ í1o \\ 
o1á \\ o1é \\ o1í \\ o1ó \\ 
u1á \\ u1ã \\ u1í \\ ú1o
}
%\emph{a-é \\ a-í \\ a-ó \\ a-ú \\ e-á \\ e-â \\ e-ã \\ e-é \\ e-ê \\ e-í \\ e-ó \\ é-o \\ e-ú \\ i-á \\ i-ã \\ í-a \\ i-é \\ i-í \\ i-ó \\ í-o \\ i-ú \\ o-á \\ o-é \\ o-í \\ o-ó \\ u-á \\ u-ã \\ u-í \\ ú-o}
\end{multicols}
See more about this on \cref{rulegrp_aa}. These rules will require exceptions for sequences with \emph{[qg]uV'} 
(where we used \emph{V'} to represent a vowel with a diacritic). See more on \cref{rulegrp_aa}.

From the list of missing hyphenations above, we find another recurring pattern, a missing hyphen preceding \emph{q}: 
% ../scripts/resultfileexamples.sh -m -f ../data/hyphenations6result-default-patterns.dic | ../scripts/filtersurrounding.sh -c '-' -n 1 | sort | uniq -c | sort -nr | awk '{print $2}' | sort | grep "q$" | sort | sed ':a;N;$!ba;s/\n/ \\\\ /g'
\begin{multicols}{5}
\setlength{\columnsep}{0pt}
\setlength{\parindent}{0pt}
\emph{a-q \\ e-q \\ i-q \\ o-q \\ r-q \\ s-q \\ u-q}
\end{multicols}
That issue might be easily solved by adding a hyphenation rule \texttt{1qu}. The \emph{q} is always followed by 
a sequence \emph{uV}, and there are already four hyphenation rules involving this sort of sequence in the default rules
(\texttt{1qu4a}, \texttt{1qu4e}, \texttt{1qu4i} and \texttt{1qu4o})\footnote{The default rules uses level three to hyphenate
between those vowels: \texttt{u3a}, \texttt{u3e}, \texttt{u3i}, and \texttt{u3o}. Apparently, they could have used 
a lower degree rule and, consequently, a lower degree rule to discourage the hyphen in sequences with \emph{q} or \emph{q}}. 
The only valid exceptions\footnote{The other exceptions we are not considering include typos and proper names (names of places, for example \emph{Urquhart}, \emph{
Qurgonteppa
}, \emph{Al-Qusayr}, etc.).} found in Wikipedia corpus are
the words \emph{far-quhar} and \emph{qu-bit}. We could then add the rule \texttt{1qu} along with the exception rules
for those hyphenation rules between \emph{u} and the following vowel (see \cref{rulegrp_aa}):
% $ grep "q" ../data/patch.TeX.pt-br.patterns | awk '{print "\\texttt{" $1 "}"}' | sed ':a;N;$!ba;s/\n/, /g'
\texttt{1qu2á}, \texttt{1qu2ã}, \texttt{1qu2â}, \texttt{1qu4é}, \texttt{1qu2ê}, and \texttt{1qu2í} (see \cref{rulegrp_aa}). 
Note that \texttt{1qu2ó} was not added since we will not use a rule to hyphenate between these vowels.
%since there are rules to create hyphenation points between \emph{u} and the following vowel (see \cref{rulegrp_aa}).  
%could be added (see \cref{rulegrp_aa,rulegrp_quo}), but they have not necessary at this moment.
%%% XXX check if they are not really necessary
% cut -d, -f1 ../data/hyphenations.csv | grep -Po "qu[aeioáéíóâêôãõ]" | sort | uniq -c | column
% 1354 qua          52 quá           8 quâ           3 quã        4411 que          99 qué          79 quê        3655 qui         224 quí         102 quo           9 quó

The missing hyphens in \emph{a-v} and \emph{i-v} might be easily solved by introducing the rule \texttt{1vô} (see \cref{rulegrp_bo}) which 
complements the rule \texttt{1vo} already included in the default set of rules.
% pi-vô, a-vô, bi*sa-vô ...
Similarly, the missing hyphens \emph{e-l}, \emph{u-l}, \emph{o-l}, and \emph{a-l} are corrected by the introduction of the rule \texttt{1lô} rule (see \cref{rulegrp_bo}),
complementing the already included rule \texttt{1lo}.
In the same way, the rules \texttt{1cô}, \texttt{1gô}, \texttt{1bô}, \texttt{1tô}, \texttt{1rô}, and \texttt{1pô} (see \cref{rulegrp_bo}) 
should be added to fix the missing hyphens \emph{e-c}, \emph{i-c}, \emph{o-g}, \emph{o-b}, \emph{a-t}, \emph{a-r},
and \emph{a-p}, respectively.
% ca*me-lô, gi*go-lô, ..., bi*be-lô, su.b-lu*nar
%                                       rule 1b2l causes wrong hyphen
%
%    .   s   u   b   l   u   n   a   r   .
%     1   0   0   |   |   |   |   |   |    1su
%     |   |   1   2   0   |   |   |   |    1b2l
%     |   |   |   1   0   0   |   |   |    1lu
%     |   |   |   |   |   1   0   0   |    1na
%max: 1   0   1   2   0   1   0   0   0
%final: s   u - b   l   u - n   a   r
The missing hyphen \emph{b-l} happened in %\emph{su.b-lu*nar}
\emph{su.b-lu*nar}, which also has a wrong hyphenation.
Those errors were caused by the default rule \texttt{1b2l}. The introduction of rule \texttt{su2b3l} 
will help solve this matter (see details in \cref{rulegrp_sub}).

% e-c, i-c: missing rule 1cô
% o-g: missing rule 1gô
% o-b: missing rule 1bô
% a-t: missing rule 1tô
% a-r: missing rule 1rô
% a-p: missing rule 1pô

The remaining missing hyphen is \emph{o-w}, which comes from the foreign \emph{kilo-watt}.
Those cases of words borrowed from other languages are dealt in the topic \ref{foreignness}.
% o-w: qui*lo-watt

% number of words with missing hyphenations using default TeX rules when analysing the words in hyphenations6.dic

@@

% cat ../data/default.TeX.pt-br.patterns <(echo -e "p2si\np2sí\n.g2no\n.g2nó\nt2c\n1p2neu\n.t2m\n.p2t\nsu2b3r\nsu2b3l") > /tmp/patts
% ./gocreateresultfile.sh ../data/hyphenations6.dic /tmp/patts > /tmp/hyphenations6result_001.dic
% ./resultfilemetrics.sh -f /tmp/hyphenations6result_001.dic
% correct: 15552; wrong: 6 (6); missing: 286 (285,1)
%
% compare with the results of the default rules:
% diff -y --suppress-common-lines <(sort ../data/hyphenations6result-default-patterns.dic) <(sort /tmp/hyphenations6result_001.dic)
% 
% ./gohyphenfull ../data/default.TeX.pt-br.patterns psiquiátrico
%
% cat ../data/default.TeX.pt-br.patterns <(echo -e ".p2si\n.p2sí\n.g2no\n.g2nó\nt2c\n1p2neu\n.t2m\n.p2t\nsu2b3r\nsu2b3l") > /tmp/patts
% ./gocreateresultfile.sh ../data/hyphenations6.dic /tmp/patts > /tmp/hyphenations6result_001.dic
% ./resultfilemetrics.sh -f /tmp/hyphenations6result_001.dic
% correct: 15562; wrong: 3 (3); missing: 279 (278,1)

After introducing the rules: \texttt{.p2si}, \texttt{.p2sí}, \texttt{.g2no}, \texttt{.g2nó}, \texttt{t2c}, 
\texttt{1p2neu}, \texttt{.t2m}, \texttt{.p2t}, \texttt{su2b3r}, and \texttt{su2b3l}, 
the number of incorrect hyphenations decreased to 3, and the number of missing hyphenations decreased to 278.
At first sight, we expected the number of incorrect hyphenations to drop to zero. However, 
the introduction of rule \texttt{su2b3l} introduced hyphenation errors in \emph{su-b.li*me}, \emph{su-b.li*ma*ção} and \emph{su-b.li*ma*do}.
%\emph{sublime}, \emph{sublimação}, and \emph{sublimado}.
%\emph{su-b.li*me}, \emph{su-b.li*ma*ção} and \emph{su-b.li*ma*do}.

\noindent\begin{minipage}{\linewidth}
\begin{lstlisting}[language={}, caption={Hyphenation of the word \emph{sublime} after the indroduction of rule \texttt{su2b3l}. 
This rule introduces a wrong hyphenation point, demanding an exception rule to fix it.}, label=sublimehyphenation]
   .   s   u   b   l   i   m   e   .
     1   0   0   |   |   |   |   |    1su
     0   0   2   3   0   |   |   |    su2b3l
     |   |   1   2   0   |   |   |    1b2l
     |   |   |   1   0   0   |   |    1li
     |   |   |   |   |   1   0   0    1me
max: 1   0   2   3   0   1   0   0
final: s   u   b - l   i - m   e
\end{lstlisting}
\end{minipage}
\noindent{}By inserting the rule \texttt{.su3b4li}, we fixed the hyphenation errors caused by the introduction of the rule \texttt{su2b3l},
as illustrated in \Cref{sublimehyphenation}. This adjustment resulted in zero hyphenation errors and 276 missing hyphens.
% cat ../data/default.TeX.pt-br.patterns <(echo -e ".p2si\n.p2sí\n.g2no\n.g2nó\nt2c\n1p2neu\n.t2m\n.p2t\nsu2b3r\nsu2b3l\n.su3b4li") > /tmp/patts
% ./gocreateresultfile.sh ../data/hyphenations6.dic /tmp/patts > /tmp/hyphenations6result_001.dic
% ./resultfilemetrics.sh -f /tmp/hyphenations6result_001.dic
% correct: 15565; wrong: 0 (); missing: 276 (275,1)

@@

The next step is to introduce the rules to address the missing hyphenation points. These rules are: 
\texttt{a1ir.}, \texttt{u1ir.}, \texttt{1qu}, \texttt{1vô}, \texttt{1lô}, \texttt{1cô}, \texttt{1gô}, \texttt{1bô}, 
\texttt{1tô}, \texttt{1rô}, \texttt{1pô}, \texttt{a1é}, \texttt{a1í}, \texttt{a1ó}, \texttt{a1ú}, \texttt{e1á}, \texttt{e1â}, \texttt{e1ã}, 
\texttt{e1é}, \texttt{e1ê}, \texttt{e1í}, \texttt{e1ó}, \texttt{é1o}, \texttt{e1ú}, \texttt{i1á}, \texttt{i1ã}, \texttt{í1a}, \texttt{i1é}, 
\texttt{i1í}, \texttt{i1ó}, \texttt{í1o}, \texttt{i1u}, \texttt{i1ú}, \texttt{o1á}, \texttt{o1é}, \texttt{o1í}, \texttt{o1ó}, \texttt{u1á}, \texttt{u1ã}, 
\texttt{u1í}, \texttt{ú1o}.
% \texttt{1qu2á}, \texttt{1qu2é}, \texttt{1qu2ê}, \texttt{1qu2í}, \texttt{que2i}, \texttt{1qu4ã}, \texttt{1qu4ó}, \texttt{1qu4â}
% \texttt{1gu2é}, \texttt{1gu2ê}, \texttt{1gu2í} 
% cat ../data/default.TeX.pt-br.patterns <(echo ".p2si, .p2sí, .g2no, .g2nó, t2c, 1p2neu, .t2m, .p2t, su2b3r, su2b3l, .su3b4li, a1ir., u1ir., 1qu, 1vô, 1lô, 1cô, 1gô, 1bô, 1tô, 1rô, 1pô, a1é, a1í, a1ó, a1ú, e1á, e1â, e1ã, e1é, e1ê, e1í, e1ó, é1o, e1ú, i1á, i1ã, í1a, i1é, i1í, i1ó, í1o, i1u, i1ú, o1á, o1é, o1í, o1ó, u1á, u1ã, u1í, ú1o" | tr -d ' ' | tr ',' '\n') > /tmp/patts2
% %%% a1â, a1ã, a1é, a1í, a1ó, a1ô, a1ú, e1á, e1â, e1ã, e1é, e1ê, e1í, e1ó, e1ô, e1ú, é1o, i1á, i1ã, i1é, i1í, i1ó, i1u, i1ú, í1a, o1á, o1ã, o1é, o1ê, o1í, o1ó, u1á, u1ã, u1é, u1ê, u1í, ú1o
% ./gocreateresultfile.sh ../data/hyphenations6.dic /tmp/patts2 > /tmp/hyphenations6result_002.dic
% ./resultfilemetrics.sh -f /tmp/hyphenations6result_002.dic
% correct: 15804; wrong: 18 (18); missing: 19 (19)
% grep "\." /tmp/hyphenations6result_002.dic | grep -oP "[^\.\*\-]{1,2}\.[^\.\*\-]{1,2}" | sort | uniq -c | sort -nr | column
% 11 qu.í          4 qu.á          1 qu.ís         1 qu.ín         1 gu.ís
After inserting these rules, the number of incorrect hyphenations increases to 18 due to the inclusion of hyphens in \emph{qu.í}, \emph{gu.í} and \emph{qu.á}. 
To overcome these errors, we need to add the exception rules: \texttt{1qu2á}, \texttt{1qu2í}, and \texttt{1gu2í}.
See more on this on \cref{rulegrp_aa}.
% cat ../data/default.TeX.pt-br.patterns <(echo ".p2si, .p2sí, .g2no, .g2nó, t2c, 1p2neu, .t2m, .p2t, su2b3r, su2b3l, .su3b4li, a1ir., u1ir., 1qu, 1vô, 1lô, 1cô, 1gô, 1bô, 1tô, 1rô, 1pô, a1é, a1í, a1ó, a1ú, e1á, e1â, e1ã, e1é, e1ê, e1í, e1ó, é1o, e1ú, i1á, i1ã, í1a, i1é, i1í, i1ó, í1o, i1u, i1ú, o1á, o1é, o1í, o1ó, u1á, u1ã, u1í, ú1o, 1qu2á, 1qu2í, 1gu2í" | tr -d ' ' | tr ',' '\n') > /tmp/patts3
% ./gocreateresultfile.sh ../data/hyphenations6.dic /tmp/patts3 > /tmp/hyphenations6result_003.dic
% ./resultfilemetrics.sh -f /tmp/hyphenations6result_003.dic
% correct: 15822; wrong: 0 (); missing: 19 (19)

We have reached a point where there is no wrong hyphenation but still there are 19 missing hyphenation. Here is the list of the missing ones:
\begin{multicols}{5}
\setlength{\columnsep}{0pt}
\setlength{\parindent}{0pt}
\emph{a-in*da \\ ra-i*nha \\ con*tri*bu-in*te \\ ra-iz \\ re-in*te*gra*ção \\ cam*pa-i*nha \\ re-in*te*grar \\ ta-i*nha \\ re-in*ci*dir \\ pa-in*ço \\ pi*xa-im \\ re-im*pres*são \\ tu-im \\ je*su-i*tis*mo \\ ben*jo-im \\ ma*lau-i*a*no \\ re-im*pri*mir}
\end{multicols}
\noindent{}Observing the instances above, 17 out of 19 involve a missing hyphen before the vowel \emph{i}. The following rules
will be used to solve those issues: \texttt{a1ind}, \texttt{a1i1nh} (see \cref{rulegrp_ain}), \texttt{e1imp} (see \cref{rulegrp_eim}), 
\texttt{e1inc}, \texttt{e1inf}, \texttt{e1ing}, \texttt{e1ins}, \texttt{e1int}, \texttt{e1inv}, (see \cref{rulegrp_ein}), 
%\texttt{u1int}, \texttt{u1ind} (see \cref{rulegrp_unit}),  (not necessary since we have the rule u1in)
and \texttt{u1iz.}, \texttt{a1iz.} (see \cref{rulegrp_aiz}).
% rulegrp_ain regra: a1ind, a1i1nh – ainda, rainha
% rulegrp_eim regra: e1imp – reimpresso
% rulegrp_ein regra: e1inc, e1inf, e1ing, e1ins, e1int, e1inv
% rulegrp_unit regra: u1int, u1ind -- contribuinte, excluindo
% rulegrp_aiz regra: u1iz., a1iz. -- juiz, raiz

% cat ../data/default.TeX.pt-br.patterns <(echo ".p2si, .p2sí, .g2no, .g2nó, t2c, 1p2neu, .t2m, .p2t, su2b3r, su2b3l, .su3b4li, a1ir., u1ir., 1qu, 1vô, 1lô, 1cô, 1gô, 1bô, 1tô, 1rô, 1pô, a1é, a1í, a1ó, a1ú, e1á, e1â, e1ã, e1é, e1ê, e1í, e1ó, é1o, e1ú, i1á, i1ã, í1a, i1é, i1í, i1ó, í1o, i1u, i1ú, o1á, o1é, o1í, o1ó, u1á, u1ã, u1í, ú1o, 1qu2á, 1qu2í, 1gu2í, a1ind, a1i1nh, e1imp, e1inc, e1inf, e1ing, e1ins, e1int, e1inv, u1int, u1ind, u1iz., a1iz." | tr -d ' ' | tr ',' '\n') > /tmp/patts4
% ./gocreateresultfile.sh ../data/hyphenations6.dic /tmp/patts4 > /tmp/hyphenations6result_004.dic
% ./resultfilemetrics.sh -f /tmp/hyphenations6result_004.dic
% correct: 15833; wrong: 0 (); missing: 8 (8)
